\documentclass[a4paper,10pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[pdftex]{hyperref}
\usepackage[pdftex]{graphicx}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{enumerate}
\usepackage{color}
\usepackage{verbatim}

\usepackage{tikz}
\usetikzlibrary{shapes,arrows}

\tikzstyle{circ} = [ circle, draw, minimum height=2em ]
\tikzstyle{line} = [draw, -latex']
\tikzstyle{none} = []

\lstset{ %
  language=C++,
  basicstyle=\footnotesize,
  numbers=left,
  numberstyle=\footnotesize,
  stepnumber=1,
}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}

\newcommand{\Domain}{\mathbb{D}}
\newcommand{\Integers}{\mathbb{Z}}
\newcommand{\Rationals}{\mathbb{Q}}
\newcommand{\Reals}{\mathbb{R}}
\newcommand{\BigO}{\mathcal{O}}

\begin{document}

\pagestyle{headings}

\title{Advanced Analysis Techniques \\ SMT Solvers}
\author{Micha\l\ Terepeta}

\date{January, 2012}

\maketitle

{\abstract The aim of this report is to provide a brief survey of current
approaches to satisfiability of Difference Logic (DL) and Unit Two Variable Per
Inequality (UTVPI) constraints. First the report covers DL with the basic
algorithms and complexity results, then goes on to describe the UTVPI case with
recent algorithms. In both cases satisfiability as well as equality generation
is covered. Furthermore the report is accompanied by a proof-of-concept
implementation of UTVPI theory solver for the Z3 SMT.}


\tableofcontents

\newpage
\section{Introduction}

In many software verification projects, it is necessary to check for
satisfiability of certain constraints. However, propositional satisfiability,
is often unsuitable for direct use in cases that involve e.g. arithmetic
constraints. This is one of the reasons for growing interest in SMT
(Satisfiability Modulo Theories). SMT solvers provide a much richer
language (or actually languages) that one can use to communicate the
constraints. The commonly available theories include free/uninterpreted
functions, equality, arrays, difference constraints, linear arithmetic, etc.

Many of these theories naturally arise in software verification, however, they
rarely occur just by themselves --- more often the constraints are defined over a
combination of various theories. This leads us to one of the main achievements
in SMT --- the Nelson-Oppen framework \cite{bib:nelson_oppen} that specifies how
one can combine various theories and still get a decidable algorithm for
satisfiability. Therefore modern SMT solvers are capable of solving the problem
of satisfiability of constraints over many different theories. This is one of
the reasons of why more and more static analyzers/software model checkers use
SMT solvers to prove (or disprove) various properties about analyzed software.

Finally there is, as always in case of software verification, the question of
performance. Today's SMT solvers use many very specialized algorithms for
various theories. As an example consider the fact that many SMT solvers have a
separate engine for linear arithmetic and for difference logic (even though
difference logic is a special case of linear arithmetic). The reason for that is
simple --- one of the best algorithms for linear real arithmetic called Simplex
is exponential in the worst case, whereas the problem of satisfiability of
linear integer arithmetic is NP-complete. On the other hand, difference
constraints can be solved by using very efficient graph-based algorithms in
polynomial time. Therefore specialization can often be quite beneficial. This
report is exactly about exploring a specialized solver for UTVPI constraints,
which are a slightly more expressive fragment of linear arithmetic yet still
allows efficient polynomial satisfiability algorithms.

%
% Difference Logic
%

\newpage
\section{Difference Logic}
\input{dl}


%
% UTVPI
%

\newpage
\section{UTVPI}
\input{utvpi}

\newpage
\section{Implementation}
\input{impl}

%
% Conclusions
%

\newpage
\section{Conclusions}

Both Difference Logic and UTVPI seem to be a great compromise for satisfiability
checking in the context of modern SMT solvers. They both allow very efficient
graph-based algorithms for problems such as satisfiability or equality
generation. And at the same time they often naturally arise in software
verification (array bounds checking, etc.) or in other problems like job
scheduling.

This report provided a brief survey of the current state-of-the-art in the
satisfiability checking for Difference Logic and UTVPI. It focused on providing
intuition about various algorithms and reasons of their efficiency. Apart from
that a proof-of-concept implementation of UTVPI theory solver for the Z3 SMT
solver has been developed.


\bibliographystyle{plain}
\bibliography{report}

\end{document}
